---
title: "Bulk RNAseq report - Carolina Scagliusi"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: united
    code_folding: hide
date: "10/21/2024"
author: "Vittoria Esposito"
---

# Overview

- RNAseq samples have been analysed with STAR/salmon using [nf-core](https://nf-co.re/rnaseq) v3.11.2

- Filtering steps: genes with reads in fewer than 11 samples are filtered out; genes with fewer reads across all samples less than nsamples/2 are filtered out.

- Differential gene expression analysis has been performed using [DESeq2](http://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html)

- Differential gene expression filters: folchange higher than log2(2) and padj < 0.01

- Gene set enrichment analysis (GSEA) has been performed using [ClusterProfiler](https://bioconductor.org/packages/release/bioc/html/clusterProfiler.html)

- GSEA filters: padj < 0.05

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r library, include=FALSE}
if (!require("DT")) install.packages('DT')

library("DESeq2")
library(rtracklayer)
library(ComplexHeatmap)
library(RColorBrewer)
library(ComplexHeatmap)
library(gplots) 
library(ggplot2)
library("DT")
library(reshape)
library(plotly)
library(clusterProfiler)
library(dplyr)
```


```{r functions, include=FALSE}
mydeseq2 = function(counts, min_reads, sample_info, min_sample = 5, ...){
  # Get counts matrix from dds object
  # counts = assay(dds)
  
  # Filter out genes with low expression
  counts = counts[rowSums(counts) > min_reads & rowSums(counts > 1) > min_sample, ]
  
  # Ensure sample_info is in the correct order matching the dds object
  sample_info <- sample_info[colnames(counts), ]
  
  ## Create new DESeq2 object accounting for batch effect
  dds_current <- DESeqDataSetFromMatrix(
    countData = counts,
    colData = sample_info,
    design = ~ batch + condition)  # Include batch effect in the design
  
  # Perform DESeq2 analysis
  diff <- DESeq(dds_current, ...)
  
  # Initialize result holders
  res_all = NULL
  df_all_genes = NULL
  
  ## Loop through conditions and perform pairwise comparisons
  for(i in 1:length(unique(sample_info$condition))){
    for(j in 1:length(unique(sample_info$condition))){
      if(i != j){
        # Extract results for condition i vs condition j
        res <- results(diff, contrast = c("condition", 
                                          as.character(unique(sample_info$condition)[i]), 
                                          as.character(unique(sample_info$condition)[j])))
        
        # Convert results to a dataframe
        res = data.frame(res)
        
        # Add comparison information
        res$comparison_n_vs_d = paste0(unique(sample_info$condition)[i], "_vs_", unique(sample_info$condition)[j])
        res$gene = rownames(res)
        
        # Append results to res_all
        res_all = rbind(res_all, res)
        
        # Filter out NA values for significant gene list
        res = na.omit(res)
        
        # Append filtered results to df_all_genes
        df_all_genes = rbind(df_all_genes, res)
      }
    }
  }
  
  ## Normalize the data using rlog (with batch effect considered)
  rld <- rlog(dds_current, blind = FALSE)
  
  # Return results as a list
  return(list(res = res_all, df_all_genes = df_all_genes, rld = rld, coldata = sample_info))
}
```

```{r parameters, echo=TRUE}

rdata_file="/Users/ieo7095/Downloads/Chiocca_group_rnaseq_various/rnaseq_carolina/results_rnaseq/star_salmon/deseq2_qc/deseq2.dds.RData"
qvalue=0.01
logfc=log2(2)
min_sample = 2 # minimum number of samples where the gene needs to have at least 1 read
org = "org.Hs.eg.db"
system("rm -r /Users/ieo7095/Downloads/Chiocca_group_rnaseq_various/rnaseq_carolina/gse_results/")
system("mkdir /Users/ieo7095/Downloads/Chiocca_group_rnaseq_various/rnaseq_carolina/gse_results")
```

```{r load_data, include=FALSE}
# load rdata from nfcore
sample_info = read.table("/Users/ieo7095/Downloads/Chiocca_group_rnaseq_various/rnaseq_carolina/samples_info.tsv", header=T)
rownames(sample_info) <- sample_info$sample
sample_info$condition <- as.factor(sample_info$condition)
sample_info$batch <- as.factor(sample_info$batch)
load(rdata_file)
```

```{r}
rownames(colData(dds)) <- sample_info$sample #change sample names
dds$sample <- sample_info$sample
# Per i campioni di controllo
controls_1 <- dds[, dds$sample == "control_1"]
controls_2 <- dds[, dds$sample == "control_2"]

# Calcolare il livello di background (media dei conteggi normalizzati)
background_1 <- rowMeans(counts(controls_1, normalized = TRUE))
background_2 <- rowMeans(counts(controls_2, normalized = TRUE))

# Campioni da 1 a 7 (HPV+ e HPV-)
hpv_samples_1_7 <- dds[, dds$sample %in% c("sample_1","sample_2","sample_3","sample_4","sample_5","sample_6","sample_7")]
counts_1_7_normalized <- counts(hpv_samples_1_7, normalized = TRUE)
counts_1_7_corrected <- round(pmax(sweep(counts_1_7_normalized, 1, background_1, FUN = "-"),0))

# Campioni da 9 a 16 (HPV+ e HPV-)
hpv_samples_9_16 <- dds[, dds$sample %in% c("sample_9","sample_10","sample_11","sample_13","sample_14","sample_15","sample_16")]
counts_9_16_normalized <- counts(hpv_samples_9_16, normalized = TRUE)
counts_9_16_corrected <- round(pmax(sweep(counts_9_16_normalized, 1, background_2, FUN = "-"),0))

# Combina i conteggi corretti per i campioni da 1 a 7 e da 9 a 16
counts_corrected_all <- cbind(counts_1_7_corrected, counts_9_16_corrected)

# Per i campioni 17-24 (già puliti, quindi usiamo i dati originali)
hpv_samples_17_25 <- dds[, dds$sample %in% c("sample_17","sample_19","sample_20","sample_21","sample_22","sample_23","sample_24","sample_25")]
counts_17_25 <- round(counts(hpv_samples_17_25, normalized = TRUE))

# Unisci tutti i dati (campioni corretti + campioni puliti)
final_counts <- cbind(counts_corrected_all, counts_17_25)

# Crea un nuovo DESeqDataSet con i conteggi corretti
dds_subset <- dds[, !dds$sample %in% c("control_1", "control_2")]
sample_info <- sample_info[sample_info$condition!="Controllo",]
# Convert sample_info to S4 DataFrame from S4Vectors
sample_info <- DataFrame(sample_info)
sample_info <- droplevels(sample_info)
# Ensure row names of sample_info match the column names of final_counts
rownames(sample_info) <- colnames(final_counts)
colData(dds_subset) <- sample_info
```


## Mild filtering

Genes total number of reads across all samples > 11 (half of the samples) are kept

```{r mild_filtering_analysis, include=FALSE}
mreads = round(ncol(dds_subset) / 2) ## filter genes without at least mreads reads in all condition combined
out = mydeseq2(counts = final_counts, min_reads = mreads, min_sample = min_sample, sample_info = sample_info)
res = out[["res"]]
rld = out[["rld"]]
df_all_genes = out[["df_all_genes"]]
coldata = out[["coldata"]]
all = na.omit(res)
```


# PCA
Let's have a look at PCA, and gene expression pattern across samples.
The batch effect has been considered in the design, but has not been corrected for this plot.
```{r mf_pca_ma, message=FALSE, warning=FALSE}
tmp = plotPCA(rld, intgroup = "condition", returnData=TRUE)
plotly::plot_ly(data=tmp, x=~PC1, y=~PC2, text = ~name, color=~condition)
```

# HPVpos vs HPVneg


### Volcano and MA plot

```{r, message=FALSE, warning=FALSE}
contrast = "HPVpos_vs_HPVneg"
res = all[all$comparison_n_vs_d == contrast, ]
select = which((res)$padj < qvalue &
                 abs((res)$log2FoldChange) > logfc)
res$significant = "nosignificant"
res$significant[select] = "significant"


plotly::plot_ly(data=res, x=~log2(baseMean), y=~log2FoldChange, color=~significant, text = ~gene, title="MA plot")

```

```{r, message=FALSE, warning=FALSE}
plotly::plot_ly(data=res, x=~log2FoldChange, y=~-log10(padj), color=~significant, text = ~gene, title="Volcano plot")
```

### Gene set enrichment analysis
The dot plot below shows the top 10 most enriched GO terms. The size of each dot correlates with the count of differentially expressed genes associated with each GO term. Furthermore, the color of each dot reflects the significance of the enrichment of the respective GO term, highlighting its relative importance.
```{r, message=FALSE, warning=FALSE, fig.height=16}
genelist = res$log2FoldChange
names(genelist) = res$gene
gse <- gseGO(geneList=sort(genelist, decreasing = T), 
                 ont = "BP", 
                 keyType = "SYMBOL", 
                 minGSSize = 3, 
                 maxGSSize = 800, 
                 pvalueCutoff = 0.05, 
                 verbose = TRUE, 
                 OrgDb = org, 
                 pAdjustMethod = "BH")

if(nrow(gse@result)>0){
  pdf(paste0("gse_results/gse.", contrast, ".BP.pdf"))
  print(dotplot(gse, showCategory=10, split=".sign") + facet_grid(.~.sign) + ggtitle("biological processes"))
  dev.off()
  
  print(dotplot(gse, showCategory=10, split=".sign") + facet_grid(.~.sign) + ggtitle("biological processes"))
  write.table(x=gse@result, file=paste0("gse_results/gse.", contrast, ".BP.tsv"), quote=F, row.names=F, sep="\t")
}

gse <- gseGO(geneList=sort(genelist, decreasing = T), 
                 ont = "CC", 
                 keyType = "SYMBOL", 
                 minGSSize = 3, 
                 maxGSSize = 800, 
                 pvalueCutoff = 0.05, 
                 verbose = TRUE, 
                 OrgDb = org, 
                 pAdjustMethod = "BH")

if(nrow(gse@result)>0){
  pdf(paste0("gse_results/gse.", contrast, ".CC.pdf"))
  print(dotplot(gse, showCategory=10, split=".sign") + facet_grid(.~.sign) + ggtitle("Cellular components"))
  dev.off()
  print(dotplot(gse, showCategory=10, split=".sign") + facet_grid(.~.sign) + ggtitle("Cellular components"))
  write.table(x=gse@result, file=paste0("gse_results/gse.", contrast, ".CC.tsv"), quote=F, row.names=F, sep="\t")
}

gse <- gseGO(geneList=sort(genelist, decreasing = T), 
                 ont = "MF", 
                 keyType = "SYMBOL", 
                 minGSSize = 3, 
                 maxGSSize = 800, 
                 pvalueCutoff = 0.05, 
                 verbose = TRUE, 
                 OrgDb = org, 
                 pAdjustMethod = "BH")
if(nrow(gse@result)>0){
  pdf(paste0("gse_results/gse.", contrast, ".MF.pdf"))
  print(dotplot(gse, showCategory=10, split=".sign") + facet_grid(.~.sign) + ggtitle("Molecular function"))
  dev.off()
  print(dotplot(gse, showCategory=10, split=".sign") + facet_grid(.~.sign) + ggtitle("Molecular function"))
  write.table(x=gse@result, file=paste0("gse_results/gse.", contrast, ".MF.tsv"), quote=F, row.names=F, sep="\t")
}
```

# Table of all genes
```{r, message=FALSE, warning=FALSE}
datatable(
  all,
  extensions = 'Buttons',
  options = list(
    dom = 'Blfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf', 'print'),
    lengthMenu = list(c(10, 25, 50, -1),
                      c(10, 25, 50, "All"))
  ),
  rownames = FALSE
)
```

# Table of all differentially expressed genes

```{r mf_table, message=FALSE, warning=FALSE}
all$FoldChange = 2^all$log2FoldChange
select = which((all)$padj < qvalue &
                 abs((all)$log2FoldChange) > logfc)
selected = all[select,]
datatable(
  selected[selected$comparison_n_vs_d == "HPVpos_vs_HPVneg", ],
  extensions = 'Buttons',
  options = list(
    dom = 'Blfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf', 'print'),
    lengthMenu = list(c(10, 25, 50, -1),
                      c(10, 25, 50, "All"))
  ),
  rownames = FALSE,
  caption = "HPVpos_vs_HPVneg"
)

all$differentially_expressed = "no"
all$differentially_expressed[select] = "yes"
final_table=selected[selected$comparison_n_vs_d == "HPVpos_vs_HPVneg",]

write.table(x=final_table, file="gse_results/differential_expression_HPVpos_vs_HPVneg.csv", quote=F, row.names=F, sep=",")
```

## HEATMAP for top 20 genes in HPVpos_vs_HPVneg
Significato dei Colori
Rosso: Indica un'espressione alta per quel gene in un determinato campione (valore sopra la media, positivo rispetto alla scala standardizzata).
Blu: Indica un'espressione bassa per quel gene in un determinato campione (valore sotto la media, negativo rispetto alla scala standardizzata).
Bianco (o colore intermedio): Indica un'espressione vicina alla media (valore standardizzato intorno a 0).
```{r}

top20_genes <- final_table %>%
  as.data.frame() %>%
  arrange(padj) %>%
  head(20) %>%
  pull(gene)

top20_counts <- all[top20_genes, ]
data_pos_vs_neg <- assay(rld)[top20_genes, ]
pheatmap(
  t(scale(t(data_pos_vs_neg))),
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  show_rownames = TRUE,
  show_colnames = TRUE,
  color = colorRampPalette(c("blue", "white", "red"))(50),
  main = "Top 20 DE Genes: HPVpos_vs_HPVneg"
)
```

